---
title: "Beta diversity"
---

This is the code for Figures 2 and 3 from *Suarez et al. 2019. Thickness determines microbial community structure and function in nitrifying biofilms via deterministic assembly. Scientific Reports. 9:5110*


#Stuff we need
```{r, message=FALSE, warning=FALSE}
library(phyloseq)
library(vegan)
library(ggplot2)
library(dplyr)
library(tidyr)
library(tibble)
```

I load the SV and tax table. SVs shorter than 200bp were excluded prior to analysis.
```{r}
UNOISE_OTU_table <- read.csv("zotutable_otu_final.csv", header=TRUE, row.names = 1)
UNOISE_tax_table <- as.matrix(read.csv("zotutable_tax_final.csv", header=TRUE, row.names = 1, na.strings=""))
metadata = read.table("metadata.txt", sep="\t", header=TRUE, row.names=1) #load the metadata

#Phyloseq stuff
d <- phyloseq(otu_table(UNOISE_OTU_table, taxa_are_rows = TRUE),
              tax_table(UNOISE_tax_table),
              sample_data(metadata))
```


#Data anaylisis
**Subsampling to even depth**
We are comparing samples with different sequencing depht, so subsampling to even depth (rarefy) might be a good idea. However some people think that subsampling is bad.
```{r}
dr <- rarefy_even_depth(physeq = d,
                         rngseed = 712)

#This function exports the OTU table to Vegan or others
veganotu = function(physeq) {
    OTU = otu_table(physeq)
    if (taxa_are_rows(OTU)) {
        OTU = t(OTU)
    }
    return(as(OTU, "matrix"))
}
OTUS.vegan <- veganotu(dr) #The subsampled OTU table outside phyloseq. I will use this later
```

##Alpha-diversity
```{r alpha_div, message=FALSE, warning=FALSE}
#I get some alpha diversity indicators with Phyloseq
Z_div = estimate_richness(dr, measures=c("Observed", "Shannon"))

Z_div = Z_div %>%
  rownames_to_column(var="names") %>%
  mutate(Carrier = sample_data(dr)$Carrier) %>% #I add the carrier type
  mutate(D1 = exp(Shannon)) %>% #True diversity (Effective number of species)
  mutate(E1 = D1/Observed) %>%
  select(-Shannon) %>%
  gather(key=Metric, value=Measurement, c("Observed", "D1", "E1"))
```

This is Figure 2A
```{r Fig_2A}
Z_div$Metric <- as.factor(Z_div$Metric)
Z_div$Metric = factor(Z_div$Metric,levels(Z_div$Metric)[c(3, 1, 2)])


alpha_names <- as_labeller(c(
  `Observed` = "Richness",
  `D1` = "Diversity",
  `E1` = "Eveness")) 

ggplot(Z_div, aes(Carrier, Measurement)) +
  facet_wrap(~ Metric, labeller = alpha_names, scales = "free") +
  geom_boxplot(aes(fill=Carrier)) +
  theme_bw() +
  theme(axis.text = element_text(colour = "black", size = 8),
        axis.title = element_blank(),
        legend.text = element_text(colour = "black", size = 8),
        legend.title = element_blank(),
      strip.background = element_blank(),
      legend.position = "none") +
  scale_fill_brewer(palette = "Set1") +
  expand_limits(y = 0)
```

**Welch T-test*** 
P-values 
 
Observed
```{r}
t.test(Measurement ~ Carrier, data = filter(Z_div, Metric=="Observed"))$p.value
```

D1
```{r}
t.test(Measurement ~ Carrier, data = filter(Z_div, Metric=="D1"))$p.value
```

E1
```{r}
t.test(Measurement ~ Carrier, data = filter(Z_div, Metric=="E1"))$p.value
```


##Beta-diversity
We use the baselga framework. We estimate **pairwise** beta-diversity, and its components.
 
Please see Baselga 2010. **Partitioning the turnover and nestedness components of beta diversity**
```{r}
library(betapart)
OTUS.vegan_bin <- as.matrix(OTUS.vegan)
OTUS.vegan_bin[OTUS.vegan_bin > 0] <- 1 #Binary

#Baselga
beta_core_bin_Z <- betapart.core(OTUS.vegan_bin)
beta_pair_bin_Z <- beta.pair(beta_core_bin_Z, index.family="sorensen")
```


Ordination - Simpson index. This is figure 2B
```{r Fig_2B}
#Source: A phyloseq object
#Distance.matrix: A distance matrix, for example Bray-curtis or Sorensen
caro.ord.PCoA <- function(Source=dr, Distance.matrix) {
  
  ordu = ordinate(Source, "PCoA", distance=Distance.matrix) #Phyloseq, makes an ordination
  ordu.data = plot_ordination(Source, ordu, justDF=TRUE) #Data for the Ordination
  
  #Here I plot the ordination
  qr <- ggplot(ordu.data, aes(Axis.1, Axis.2)) +
  geom_point(size = 3, 
             alpha=0.7, 
             aes(shape=factor(Carrier), #My environmental variable is carrier type
                 fill=factor(Carrier)), 
             stroke=0.5) +
  scale_shape_manual(values = c(21, 24)) + #I want shapes also
  theme_bw() +
  theme(panel.border = element_rect(colour = "black"),
        aspect.ratio=1, 
        axis.ticks = element_blank(),
        panel.grid = element_blank(),
        axis.text = element_text(size =7, color = "black"), 
        axis.title = element_blank(), 
        axis.text.x = element_blank(), 
        axis.text.y = element_blank(),
        legend.title = element_blank(),
        legend.text = element_text(size =8, color = "black"),
       legend.position="right") +
  scale_fill_brewer(type = "qual", palette = "Set1") #I love colorbrewer
  return(qr)
}

#Figure 2B
caro.ord.PCoA(dr, beta_pair_bin_Z$beta.sor)
```
 
  
##Figure 3
```{r}
#We convert the dist to a matrix
DIST_TO_MATRIX <- function(Source, Type) {
  #@Source = a dist object
  #@Type = a description of the dist matrix
  library(reshape2)
  RC_mod <- as.matrix(Source)
  RC_mod <- melt(RC_mod)[melt(lower.tri(RC_mod))$value,]
  names(RC_mod) <- c("c1", "c2", "distance")
  qr = RC_mod %>%
    mutate(Carrier=paste(substr(c1, 1, 3), substr(c2, 1, 3), sep="_"))
  qr$type <- Type 
  return(qr)
  detach("package:reshape2", unload=TRUE)
}

beta_pair_bin_Z.sor <- DIST_TO_MATRIX(beta_pair_bin_Z$beta.sor, "Total")
beta_pair_bin_Z.sne <- DIST_TO_MATRIX(beta_pair_bin_Z$beta.sne, "Nestedness")
beta_pair_bin_Z.sim <- DIST_TO_MATRIX(beta_pair_bin_Z$beta.sim, "Turnover")
```
  
 
This is **figure 3B**
```{r Fig_3B}
#We put all together
beta_pair_bin_Z.all <- beta_pair_bin_Z.sor %>%
  bind_rows(beta_pair_bin_Z.sne) %>%
  bind_rows(beta_pair_bin_Z.sim)

#Reorder levels
#pairwise comparisons
beta_pair_bin_Z.all$Carrier <- as.factor(beta_pair_bin_Z.all$Carrier)
beta_pair_bin_Z.all$Carrier = factor(beta_pair_bin_Z.all$Carrier,levels(beta_pair_bin_Z.all$Carrier)[c(2,1,3)])
#Beta partition
beta_pair_bin_Z.all$type <- as.factor(beta_pair_bin_Z.all$type)
beta_pair_bin_Z.all$type = factor(beta_pair_bin_Z.all$type,levels(beta_pair_bin_Z.all$type)[c(2,3,1)])


ggplot(beta_pair_bin_Z.all, aes(y=distance, x=type)) +
  geom_boxplot(aes(fill=Carrier)) +
  theme_bw() +
  theme(axis.text = element_text(colour = "black", size = 8),
        axis.title = element_text(colour = "black", size = 8),
        legend.text = element_text(colour = "black", size = 8),
        legend.title = element_blank(),
      strip.background = element_blank(),
      #axis.text.x=element_blank(),
      axis.title.x=element_blank(),
      legend.position = "none") +
  scale_color_brewer(palette = "Set1") +
  labs(y="Pairwise diversity", fill="Carrier") +
  scale_x_discrete(labels=c(expression(beta*"sor"), 
                            expression(beta*"sim"), 
                            expression(beta*"sne"))) +
  #scale_x_discrete("Carrier type pairwise comparisons") +
  scale_fill_brewer(palette = "Set1", labels=c("within Z400", "within Z50", "between-group")) 
```
   
 
Beta diversity ratio: **Figure 3C**
```{r Fig_3C}
beta_pair_bin_Z.com <- dplyr::rename(beta_pair_bin_Z.sor, sorensen = distance) 
beta_pair_bin_Z.com$nestedness <- beta_pair_bin_Z.sne$distance
beta_pair_bin_Z.com <- beta_pair_bin_Z.com  %>%
  mutate(B.ratio = nestedness/sorensen)

#Reorder levels
#pairwise comparisons
beta_pair_bin_Z.com$Carrier <- as.factor(beta_pair_bin_Z.com$Carrier)
beta_pair_bin_Z.com$Carrier = factor(beta_pair_bin_Z.com$Carrier,levels(beta_pair_bin_Z.com$Carrier)[c(2,1,3)])


ggplot(beta_pair_bin_Z.com, aes(y=B.ratio, x=Carrier)) +
  geom_boxplot(aes(fill=Carrier)) +
  #geom_jitter(width = 0.25, height = 0, size=2, fill="white", alpha=0.5) +
  theme_bw() +
theme(axis.text = element_text(colour = "black", size = 8),
        axis.title = element_text(colour = "black", size = 8),
        legend.text = element_text(colour = "black", size = 8),
        legend.title = element_blank(),
      strip.background = element_blank(),
      axis.text.x=element_blank(),
      axis.title.x=element_blank(),
      legend.position = "right") +
  scale_fill_brewer(palette = "Set1", labels=c("between-group", 
                                               "within Z400", 
                                               "within Z50")) +
  labs(y= expression(beta*"ratio"), fill="Carrier")
```

###PERMANOVA
Sorensen
```{r}
adonis(beta_pair_bin_Z$beta.sor ~ Carrier, metadata, perm=999)
```

Simpson
```{r}
adonis(beta_pair_bin_Z$beta.sim ~ Carrier, metadata, perm=999)
```

Beta-diversity due to nestedness
```{r}
adonis(beta_pair_bin_Z$beta.sne ~ Carrier, metadata, perm=999)
```


###Standarized size effect of beta diversity
I use a model with fixed rows and columns (Fixed richness and incidence). the algorithm I use is quasiswap. This might not be the *best* null model, nor the *best* algorithm. It might be too conservative, and fixed-fixed models have some issues. Use with caution.
  
**Important!** To recreate the plots from Suarez et al (2019) use nsimul = 999
```{r}
#Regional species pool
sordist <- function(x) vegdist(x, binary=TRUE)
#quasiswap might be better than swap.
#thin can reduce the bias of null models.

#Use nsimul = 999
z_null_sor <- oecosimu(OTUS.vegan, sordist, method = "quasiswap", nsimul = 9, thin=1000)
```

Using local species pool for within comparisons. Here I look at **Z50** carriers
```{r}
d.z50 = subset_samples(d, Carrier=="Z50")
OTUS.vegan.z50 = veganotu(d.z50)
z_null_sor.z50 <- oecosimu(OTUS.vegan.z50, sordist, method = "quasiswap", nsimul = 9, thin=1000)
```

Using local species pool for within-group comparisons.Here I look at **Z400** carriers
```{r}
d.z400 = subset_samples(d, Carrier=="Z400")
OTUS.vegan.z400 = veganotu(d.z400)
z_null_sor.z400 <- oecosimu(OTUS.vegan.z400, sordist, method = "quasiswap", nsimul = 9, thin=1000)
```

This is Figure 3A. With local species pool for within-group comparisons. Regional pool for between-group comparison
```{r, warning=FALSE}
#I want the Z-values (SES)
beta_pair_z_null_sor.z50 <- DIST_TO_MATRIX(z_null_sor.z50$oecosimu$z, "within Z50")
beta_pair_z_null_sor.z400 <- DIST_TO_MATRIX(z_null_sor.z400$oecosimu$z, "within Z400")
beta_pair_z_null_sor <- DIST_TO_MATRIX(z_null_sor$oecosimu$z, "between-group") %>%
  filter(Carrier == "Z40_Z50")

beta_pair_bin_Z.all <- beta_pair_z_null_sor %>%
  bind_rows(beta_pair_z_null_sor.z50) %>%
  bind_rows(beta_pair_z_null_sor.z400)

ggplot(beta_pair_bin_Z.all, aes(x=type, y=distance, fill=type)) +
  geom_hline(yintercept = -2, color="black", linetype = 2) +
  geom_hline(yintercept =  2, color="black", linetype = 2) +
  geom_boxplot() +
  theme_bw() +
 theme(axis.text = element_text(colour = "black", size = 8),
        axis.title = element_text(colour = "black", size = 8),
        legend.text = element_text(colour = "black", size = 8),
        legend.title = element_blank(),
      strip.background = element_blank(),
      axis.text.x=element_blank(),
      axis.title.x=element_blank(),
      legend.position = "right") +
  labs(y="Standardized effect size S?rensen", fill="type") +
  scale_fill_brewer(palette = "Set1")
```



